Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-05-21 doc/thesis/intro.tex
Most of Alex's changes done

In evolutionary biology, a \defn{phylogeny}, or \defn{phylogenetic tree}, is a
graphical representation of the the evolutionary relationships among a group of
organisms or species (generally, \defn{taxa})~\autocite{haeckel1866generelle}.
The \defn{tips} of a phylogeny, that is, the nodes without any descendants,
correspond to \defn{extant}, or observed, taxa. The \defn{internal nodes}
correspond to their (usually extinct) common ancestors. The edges or
\defn{branches} of the phylogeny connect ancestors to their descendants.
Phylogenies may have a \defn{root}, which is a node with no descendants
distinguished as the most recent common ancestor of all the extant
taxa~\autocite{harding1971probabilities}. When such a root exists, the tree is
referred to as being \defn{rooted}; otherwise, it is \defn{unrooted}. The
structural arrangement of nodes and edges in the tree is referred to as its
\defn{topology}~\autocite{cavalli1967phylogenetic}.

@@

The kernel score of a pair of trees, denoted $K(T_1, T_2)$, is defined as a sum
over all pairs of nodes $(n_1, n_2)$, where $n_1$ is a node in $T_1$ and $n_2$
is a node in $T_2$. Following \textcite{poon2013mapping}, let $N(T)$ denote the
set of all nodes in $T$, $\nc(n)$ be the number of children of node $n$,
$c_n^{j}$ be the $j$th child of node $n$, and $l_n$ be the vector of branch
lengths connecting node $n$ to its descendants. The \defn{production rule} of
$n$ is its total number of children, and its number of leaf children. That is,
if two nodes have the same number of children and among these, the same number
of leaves, then they have the same production rule. Let $k_G(x, y)$ be a
Gaussian radial basis function of the vectors $x$ and $y$,
\[
  k_G(x, y) = \exp\left(-\frac{1}{2\sigma} \norm{x - y}_2^2\right),
\]
where $\norm{\cdot}_2$ is the Euclidean norm and $\sigma$ is a variance
parameter. The tree kernel is defined as 
\[
  K(T_1, T_2) = \sum_{n_1 \in N(T_1)} \sum{n_2 \in N(T_2)} \Delta (n_1, n_2),
\]
where
\[
  \Delta(n_1, n_2) =
  \begin{cases}
    \lambda & n_1 \text{ and } n_2 \text{ are leaves} \\
    \lambda k_G(l_{n_1}, l_{n_2}) \displaystyle\prod_{j=1}^{\nc(n_1)} \left(1 +
    \Delta(c_{n_1}^j, c_{n_2}^j) \right) & \begin{aligned} n_1 \text{ and } n_2 \text{ have the same} \\ \text{production rule} \end{aligned} \\
    0 & \text{otherwise}.

  \end{cases}
\]
Here $\lambda$ is a decay factor parameter, which penalizes large matches that
tend to dominate the kernel score. In this work, we refer to the parameters
$\lambda$ and $\sigma$ as \defn{meta-parameters}, to avoid confusing them with
model parameters we are trying to estimate.

%The tree kernel was later shown to be highly effective in differentiating trees
%simulated under a compartmental model with two risk groups of varying contact
%rates~\autocite{poon2015phylodynamic}. In that paper,
%\citeauthor{poon2015phylodynamic} used the tree kernel as the distance function
%in \gls{ABC} (see \cref{sec:abc}), to fit epidemiological models to observed
%trees. This is an example of kernel-ABC~\autocite{nakagome2013kernel}, which
%will be discussed further in \cref{sec:abc}.

@@

\Gls{SMC} is the name for a family of statistical inference methods which rely
on approximating probability distributions of interest with large collections
of \defn{particles}, here denoted
$\set{x^{(k)}}$~\autocite{doucet2001introduction, liu2008monte}. These
collections or \defn{populations} are constructed to form a \defn{Monte Carlo
approximation} to some distribution of interest $\pi$, meaning that the
empirical distribution of the particles converges in distribution to $\pi$ as
the population size gets large~\autocite{liu2001theoretical}. The word
\defn{sequential} is used because the particle populations are modified in an
iterative fashion over time, for example, to incorporate new evidence.

@@

\subsection{Sequential importance sampling}
\label{subsec:sis}

\Gls{SIS}~\autocite{gordon1993novel} is one type of \gls{SMC} method, whose aim
is to sample from a distribution $\pi$ on an high-dimensional space, say
$\pi(\vec{x}) = \pi(x_1, \ldots, x_d)$. The basis of \gls{SIS} is \gls{IS},
which is a method of estimating summary statistics of distributions which are
known only up to a normalizing constant, and therefore cannot be sampled from
directly. That is, if $\pi$ is such a distribution and $f$ is any real-valued
function, \gls{IS} is concerned with estimating
\[
  \pi(f) = \int f(x)\pi(x)\d x = \int f(x) \frac{\gamma(x)}{Z} \d x,
\]
where the integral is over the space on which $\pi$ is defined, $\gamma(x)$ is
known pointwise, and $Z = \int \gamma(x) \d x$ is the unknown normalizing
constant. Suppose we have at hand another distribution $\eta$, called the
\defn{importance distribution}, from which we are able to sample. Define the
\defn{importance weight} as the ratio $w(x) = \gamma(x)/\eta(x)$. We can
express the normalizing constant $Z$ in terms of the importance weight and
distribution, $Z = \int w(x) \eta(x) \d x$, and in turn write the expectation
of interest as
\[
  \int f(x) \pi(x) \d x = \frac{\int f(x) \gamma(x) \d x}
                               {\int w(x) \eta(x) \d x}.
\]
If we sample a large number of points from $\eta$, then $\eta(x)$ can be
approximated by a Monte Carlo estimate. Since the remaining quantities $f$,
$\gamma$, and $w$ can all be evaluated pointwise, these are all the ingredients
we need to obtain an estimate of $\pi(f)$. Although this is a simple and
elegant approach, the drawback is that the variance of the estimate is
proportional to the variance of the importance weights~\autocite{liu2008monte},
which may be quite large if $\eta$ and $\gamma$ are very different. Therefore,
the practical use of \gls{IS} on its own is limited, since it depends on
finding an importance distribution similar to $\pi$, which we usually know very
little about \textit{a priori}.

@@

The recursive definition \cref{eq:sisw} suggests an algorithm for obtaining a
sample from $\pi$ (\cref{alg:sis}).  We begin with $n$ ``particles'' which have
been sampled from the importance distribution $\eta(x_0)$ for $\pi(x_0)$. The
particles are updated and reweighted $d$ times, corresponding to the $d$
elements of the decomposition of $\pi$. At the $i$th step, each particle is
extended to include $x_i$ drawn according to the chosen $\eta(x_i \mid
\vec{x_{i-1}})$, and the importance weights are recalculated and normalized.

\begin{algorithm}
  \caption{Sequential importance sampling.}
  \begin{algorithmic}
    \For {$k = 1$ to $n$}
      \State Sample $x_1^{(k)}$ from $\eta(x_1)$
      \Comment{Initialize the $k$th particle}
      \State $w^{(k)} \gets \dfrac{\pi\left(x_1^{(k)}\right)}{\eta\left(x_1^{(k)}\right)}$
    \EndFor
    \For {$i = 2$ to $d$}
      \For {$k = 1$ to $n$}
        \State Sample $x_i^{(k)}$ from $\eta\left(x_i \mid \vec{x_{i-1}^{(k)}}\right)$
        \Comment Extend the $k$th particle
        \State $w(\vec{x_i}^{(k)}) \gets \dfrac{\pi\left(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}\right)}{\eta\left(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}\right)} \cdot w(\vec{x_{i-1}}^{(k)})$
      \EndFor
      \State Normalize the weights so that $\sum w = 1$
    \EndFor
    \State Sample $n$ particles with probabilities $w$
  \end{algorithmic}
  \label{alg:sis}
\end{algorithm}

Of course, $\eta$ is merely an approximation to $\pi$, and may be a fairly poor
one depending on the application. Try as we might to keep the variances of the
weights low, the cumulative errors at each sequential step tend to push many of
the weights to very low values. This results in a poor approximation to $\pi$,
since only a few particles retain high importance weights after all $d$
sequential steps. To mitigate this problem, a resampling step is periodically
applied when the variance in the importance weights becomes too high. Several
different criteria have been proposed for when to resample, but we focus here
on the one described by \textcite{liu2008monte}, namely the decay of the
\gls{ESS} below a prescribed threshold, conventionally $n/2$. The \gls{ESS} of
the population of particles is defined as \[ \ESS(w) = \frac{n}{1 + \Var(w)},
\] where $n$ is the number of particles~\autocite{liu2008monte}. When the
\gls{ESS} drops below the threshold, particles are resampled according to their
weights. This results in the removal of low-weight particles from the
population, and also equalizes all the weights. Various resampling strategies
beyond the basic sampling with replacement have been
proposed~\autocite{douc2005comparison}, but we will not discuss those here.

@@

The \gls{SIS} algorithm described above aims to sample from a high-dimensional
distribution $\pi(x)$, by sequentially sampling from $d$ distributions of lower
but increasing dimension. \textcite{del2006sequential} developed an
\defn{\gls{SMC} sampler} with an alternative objective: to sample sequentially
from $d$ distributions $\pi_1, \ldots, \pi_d$, all of the same dimension and
defined on the same space. The $\pi_i$ are assumed to form a related sequence,
such as posterior distributions attained by sequentially considering new
evidence. As with \gls{SIS}, we assume that $\pi_i(x) = \gamma_i(x) / Z_i$,
where $\gamma_i$ is known pointwise and the normalizing constant $Z_i$ is
unknown.

@@

\begin{algorithm}
  \caption{Sequential Monte Carlo sampler of \textcite{del2006sequential}.}
  \begin{algorithmic}
    \For {$k = 1$ to $n$}
      \State Sample $x_1^{(k)}$ from $\eta_1(x_1)$
      \Comment{Initialize the $k$th particle}
      \State $w^{(k)} \gets \dfrac{\gamma_1\left(x_1^{(k)}\right)}{\eta_1\left(x_1^{(k)}\right)}$
      \State Normalize the weights so that $\sum w = 1$
    \EndFor
    \For {$i = 2$ to $d$}
      \For {$k = 1$ to $n$}
        \State Sample $x_i^{(k)}$ from $K(x_{i-1}^{(k)}, x_i)$
        \Comment Extend the $k$th particle
        \State $w^{(k)} \gets w^{(k)} \cdot \dfrac{\gamma_i(x_i) L_{i-1}(x_i, x_{i-1})}{\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}$
      \EndFor
      \State Normalize the weights so that $\sum w = 1$
      \If {$\ESS(w) < T$}
        \Comment{$T$ is a user-defined threshold}
        \State Resample the particles according to $w$
        \For {$k = 1$ to $n$}
          \State $w^{(k)} \gets 1/n$
        \EndFor
      \EndIf
      \State Sample the $i$th component of $n$ particles with probabilities $w$
    \EndFor
  \end{algorithmic}
  \label{alg:smcsamp}
\end{algorithm}

@@

For a general model, the probability density of $y$ given the parameters
$\theta$ is also known as the \defn{likelihood}, written $\L$, of $\theta$.
That is, $\L(\theta \mid y) = f(y \mid \theta)$ for the model's \gls{pdf} $f$.
The higher the value of the likelihood, the more likely the observations $y$
are under the model. Thus, the likelihood provides a natural criterion for
fitting the model parameters: we want to pick $\theta$ such that the
probability density of our observed outcomes $y$ is as high as possible. The
parameters which optimize the likelihood are known as the \textit{\gls{ML}}
estimates, denoted $\hat{\theta}$. That is,
\[
  \hat{\theta} = \argmax_\theta\; \L(\theta \mid y).
\]
\Gls{ML} estimation is usually performed with numerical optimization. In the
simplest terms, many possible values for $\theta$ are examined, $\L(\theta \mid
y)$ is calculated for each, and the parameters which produce the highest value
are accepted. Many sophisticated numerical optimization methods exist, although
they may not be guaranteed to find the true \gls{ML} estimates if the
likelihood function is complex.

@@

Ordinarily, Bayesian inference targets the posterior distribution $f(\theta
\mid y)$. That is, in the Bayesian framework, model parameters with higher
posterior density are ``better'' in the sense that they offer a more credible
explanation for the observed data. Approximate Bayesian computation offers an
alternative metric for parameter credibility, namely the similarity of
simulated datasets to the observed data. If datasets simulated under the model
closely resemble the real data, it follows that the model is a reasonable
approximation to the real-world process generating the observed data. More
formally, suppose we have a distance measure $\rho$ defined on the space of all
possible data our model could generate. \gls{ABC} aims to sample from the joint
posterior distribution of model parameters and simulated datasets $z$ which are
within some small distance $\varepsilon$ of the observed data $y$,
\[
  \pi_{\varepsilon}(\theta, z \mid y) =
  \frac{\pi(\theta) f(z \mid \theta) \I_{A_{\varepsilon, y}} (z)}
  {\int_{A_{\varepsilon, y} \times \Theta} \pi(\theta) f(z \mid \theta) \d \theta}.
\]
Here, $A_{\varepsilon, y}$ is an $\varepsilon$-ball around $y$ with
respect to $\rho$, $\Theta$ is the space of all possible model parameters, and
$\I$ is the indicator function~\autocite{marin2012approximate}. As we shall
see in the next section, this distribution can be sampled from exactly. The
word ``approximate'' derives from the assumption that, for a suitably chosen
distance $\rho$ and a small enough $\varepsilon$, the marginal in $z$ of
this distribution approximates the posterior of
interest~\autocite{marin2012approximate}. That is,
\[
  \int \pi_\varepsilon(\theta, z \mid y) \d z \approx f(\theta \mid y).
\]
The intuition for why this approximation might hold comes from the fact that, 
when $\varepsilon = 0$, the integral on the left is exactly equal to the
posterior. Thus, by taking $\varepsilon$ small, we should attain something
close to the posterior.

The distribution $\pi_\varepsilon(\theta, z \mid y)$ is variously referred to
as the  \textit{\gls{ABC} target distribution} or the \gls{ABC} approximation
to the posterior. Note that in many formulations, the distance function $\rho$
is defined as $\rho(S(\cdot), S(\cdot))$ where $S$ is a function which maps
data points into a vector of summary statistics. This can be useful if the data
are high-dimensional or of a complex type, but it is not strictly necessary.
For instance, if the data are numeric and of low dimension, the distance
function may simply be the Euclidean distance~\autocite{sisson2007sequential}.
For more complex data, \textcite{nakagome2013kernel} proposed the use of a
kernel function (defined in \cref{subsec:treeshape}), an approach they dubbed
\defn{kernel-\gls{ABC}}.

@@

Algorithms for performing \gls{ABC} fall into one of three categories:
rejection, \gls{MCMC}, and \gls{SMC}~\autocite{marin2012approximate}. To
simplify the math, we shall restrict the descriptions of these algorithms to
the case of one simulated dataset per parameter particle (the meaning of this
will become clear shortly). The extension to multiple datasets per particle is
straightforward and will be given at the end of the section. We use the
variable $x$ to refer to the pair $(\theta, z)$, so that the \gls{ABC} target
distribution can be written $\pi_\varepsilon(x \mid y)$.

>>>>

In evolutionary biology, a \defn{phylogeny}, or \defn{phylogenetic tree}, is a
graphical representation of the the evolutionary relationships among a group of
organisms or species (generally, \defn{taxa})~\autocite{haeckel1866generelle}.
The \defn{tips} of a phylogeny, that is, the nodes without any descendants,
correspond to \defn{extant}, or observed, taxa. The \defn{internal nodes}
correspond to their common ancestors, {\color{blue}\uline{usually extinct
(although occasionally the internal nodes may be observed as well, eg.
\autocite{jombart2011reconstructing}).}} The edges or \defn{branches} of the
phylogeny connect ancestors to their descendants. Phylogenies may have a
\defn{root}, which is a node with no descendants distinguished as the most
recent common ancestor of all the extant
taxa~\autocite{harding1971probabilities}. When such a root exists, the tree is
referred to as being \defn{rooted}; otherwise, it is \defn{unrooted}. The
structural arrangement of nodes and edges in the tree is referred to as its
\defn{topology}~\autocite{cavalli1967phylogenetic}.

@@

The kernel score of a pair of trees, denoted $K(T_1, T_2)$, is defined as a sum
over all pairs of nodes $(n_1, n_2)$, where $n_1$ is a node in $T_1$ and $n_2$
is a node in $T_2$. Following \textcite{poon2013mapping}, let $N(T)$ denote the
set of all nodes in $T$, $\nc(n)$ be the number of children of node $n$,
$c_n^{j}$ be the $j$th child of node $n$, and $l_n$ be the vector of branch
lengths connecting node $n$ to its descendants. The \defn{production rule} of
$n$ is its total number of children, and its number of leaf children. That is,
if two nodes have the same number of children and among these, the same number
of leaves, then they have the same production rule. Let $k_G(x, y)$ be a
Gaussian radial basis function of the vectors $x$ and $y$,
\[
  k_G(x, y) = \exp\left(-\frac{1}{2\sigma} \norm{x - y}_2^2\right),
\]
where $\norm{\cdot}_2$ is the Euclidean norm and $\sigma$ is a variance
parameter. The tree kernel is defined as 
\begin{align}
  K(T_1, T_2) = \sum_{n_1 \in N(T_1)} \sum_{n_2 \in N(T_2)} \Delta (n_1, n_2),
  \label{eq:kernel}
\end{align}
where
\[
  \Delta(n_1, n_2) =
  \begin{cases}
    \lambda & n_1 \text{ and } n_2 \text{ are leaves} \\
    \lambda k_G(l_{n_1}, l_{n_2}) \displaystyle\prod_{j=1}^{\nc(n_1)} \left(1 +
    \Delta(c_{n_1}^j, c_{n_2}^j) \right) & \begin{aligned} n_1 \text{ and } n_2 \text{ have the same} \\ \text{production rule} \end{aligned} \\
    0 & \text{otherwise}.
  \end{cases}
\]
{\color{blue}\uline{
The parameter $\lambda$ in the above expression, is called the \emph{decay
factor}~\autocite{moschitti2006making}, and takes a value between 0 and 1.
Without this parameter, terms in the sum} \ref{eq:kernel} \uline{corresponding
to large subset trees with the same topology would be similarly large and tend
to dominate the kernel score. $\lambda$ penalizes $\Delta$ more strongly as the
number of recursive calls increases, which downweights the largest matching
substructures and allows smaller matches to contribute more to the kernel
score. }} In this work, we refer to the parameters $\lambda$ and $\sigma$ as
\defn{meta-parameters}, to avoid confusing them with model parameters we are
trying to estimate. {\color{blue}\uline{ When evaluating the tree kernel, it is
helpful to reorder the children of each internal node such that the larger of
the two subtrees is on the right-hand side. If the two subtrees have equal
sizes, then the child with the longer branch length can be put on the
right-hand side. This operation is referred to as \defn{ladderizing}. Since the
ordering of children is arbitrary in phylogenies, this operation ensures that a
maximal number of matching subset trees are counted by the tree kernel without 
making meaningful changes to the trees. }}

{\color{blue}\uline{
The tree kernel was later shown to be highly effective in differentiating trees
simulated under a compartmental model with two risk groups of varying contact
rates~\autocite{poon2015phylodynamic}. In that paper,
\citeauthor{poon2015phylodynamic} used the tree kernel as the distance function
in \gls{ABC}} (see \cref{sec:abc}), \uline{to fit epidemiological models to
observed trees. }}

@@

\Gls{SMC} is the name for a family of statistical inference methods which rely
on approximating probability distributions of interest with large collections
of \defn{particles}, here denoted
$\set{x^{(k)}}$~\autocite{doucet2000sequential, doucet2001introduction,
liu2008monte}. These collections or \defn{populations} are constructed to form
a \defn{Monte Carlo approximation} to some distribution of interest $\pi$,
meaning that the empirical distribution of the particles converges in
distribution to $\pi$ as the population size gets
large~\autocite{liu2001theoretical}. The word \defn{sequential} is used because
the particle populations are modified in an iterative fashion over time, for
example, to incorporate new evidence.

@@

{\color{blue}\uline{
The goal of this section of the introduction is to describe an algorithm called
the \gls{SMC} sampler~\autocite{del2006sequential}, which forms the basis of
the adaptive \gls{ABC}-\gls{SMC} algorithm we apply toward the main objective
of this work. We begin by describing \gls{SIS}, which is a precursor to
\gls{SMC} that samples from a sequence of distributions defined on spaces of
increasing dimension. We then describe \gls{SMC} itself, which extends
\gls{SIS} with a resampling step to fight particle degeneracy. Finally, we
outline the \gls{SMC} sampler, which allows \gls{SMC} to be applied to
sequences of distributions all defined on the same space. This terminology will
become clear as the methods are described.
}}

\subsection{Sequential importance sampling}
\label{subsec:sis}
\glsreset{SIS}

\Gls{SIS}~\autocite{gordon1993novel} {\color{blue}\uline{is a particle-based
method}} whose aim is to sample from a distribution $\pi$ on an
high-dimensional space, say $\pi(\vec{x}) = \pi(x_1, \ldots, x_d)$. The basis
of \gls{SIS} is \gls{IS}, which is a method of estimating summary statistics of
distributions which are known only up to a normalizing constant, and therefore
cannot be sampled from directly. That is, if $\pi$ is such a distribution and
$f$ is any real-valued function, \gls{IS} is concerned with estimating
\[
  \pi(f) = \int f(x)\pi(x)\d x = \int f(x) \frac{\gamma(x)}{Z} \d x,
\]
where the integral is over the space on which $\pi$ is defined, $\gamma(x)$ is
known pointwise, and $Z = \int \gamma(x) \d x$ is the unknown normalizing
constant. Suppose we have at hand another distribution $\eta$, called the
\defn{importance distribution}, from which we are able to sample. Define the
\defn{importance weight} as the ratio $w(x) = \gamma(x)/\eta(x)$. We can write
the expectation of interest as
{\color{blue}
\begin{align}
  \int f(x) \pi(x) \d x = \frac{1}{Z} \int w(x) f(x) \d x.
  \label{eq:is}
\end{align}
}
{\color{blue}\uline{
Since $\eta$ can be sampled from exactly, and $\gamma$ and $f$ can both be
evaluated pointwise, the integral $\int w(x) f(x) \d x$ can be approximated by
a Monte Carlo estimate. Moreover, the normalizing constant $Z$ can be expressed
in terms of the importance weight and distribution, $Z = \int w(x) \eta(x) \d
x$. Therefore, we have all the ingredients we need to obtain an estimate of
$\pi(f)$ using} \cref{eq:is}.} Although this is a simple and elegant approach,
the drawback is that the variance of the estimate is proportional to the
variance of the importance weights~\autocite{liu2008monte}, which may be quite
large if $\eta$ and $\gamma$ are very different. Therefore, the practical use
of \gls{IS} on its own is limited, since it depends on finding an importance
distribution similar to $\pi$, which we usually know very little about
\textit{a priori}.

@@

The recursive definition \ref{eq:sisw} suggests an algorithm for obtaining a
sample from {\color{blue}\uline{$\eta$ and using it to obtain an approximate
sample from $\pi$ by \gls{IS}}} (\cref{alg:sis}).  We begin with $n$
``particles'' which have been sampled from the importance distribution
$\eta(x_0)$ for $\pi(x_0)$. The particles are updated and reweighted $d$ times,
corresponding to the $d$ elements of the decomposition of $\pi$. At the $i$th
step, each particle is extended to include $x_i$ drawn according to the chosen
$\eta(x_i \mid \vec{x_{i-1}})$, and the importance weights are recalculated and
normalized.

\begin{algorithm}
  \caption{Sequential importance sampling.}
  \begin{algorithmic}
    \For {$k = 1$ to $n$}
      \State Sample $x_1^{(k)}$ from $\eta(x_1)$
      \Comment{Initialize the $k$th particle}
      %\State $w^{(k)} \gets \dfrac{\pi\left(x_1^{(k)}\right)}{\eta\left(x_1^{(k)}\right)}$
      \State $w^{(k)} \gets \pi(x_1^{(k)}) / \eta(x_1^{(k)})$
    \EndFor
    \For {$i = 2$ to $d$}
      \For {$k = 1$ to $n$}
        \State Sample $x_i^{(k)}$ from $\eta(x_i \mid \vec{x_{i-1}^{(k)}})$
        \Comment Extend the $k$th particle
        \State $w(\vec{x_i}^{(k)}) \gets [\pi(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}) \,/\, \eta(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}})] \cdot w(\vec{x_{i-1}}^{(k)})$
      \EndFor
      \State Normalize the weights so that $\sum w = 1$
    \EndFor
  \end{algorithmic}
  \label{alg:sis}
\end{algorithm}

\subsection{Sequential Monte Carlo}

The importance distribution $\eta$ constructed with \gls{SIS} is merely an
approximation to $\pi$, and may be a fairly poor one in practice depending on
the application. Try as we might to keep the variances of the weights low, the
cumulative errors at each sequential step tend to push many of the weights to
very low values~\autocite{doucet2000sequential}. This results in a poor
approximation to $\pi$, since only a few particles retain high importance
weights after all $d$ sequential steps, 
{\color{blue}\uline{ a problem known as \defn{particle degeneracy}. To mitigate
this problem, \textcite{gordon1993novel} introduced technique they called the
\defn{bootstrap filter}, which involves a resampling of the population of
particles after each sequential step in accordance with their importance
weights. A similar idea, termed \defn{particle rejuvination}, was proposed by
\textcite{liu1995blind}. These approaches cause particles with high importance
weights to be replicated in the population, while particles with low weights
may be removed. After each resampling step, the importance weights for all
particles are set equal.}}

\begin{algorithm}
  \caption{Sequential Monte Carlo \autocite{doucet2000sequential}.}
  \begin{algorithmic}
    \For {$k = 1$ to $n$}
      \State Sample $x_1^{(k)}$ from $\eta(x_1)$
      \Comment{Initialize the $k$th particle}
      \State $w^{(k)} \gets \pi(x_1^{(k)}) / \eta(x_1^{(k)})$
    \EndFor
    \For {$i = 2$ to $d$}
      \For {$k = 1$ to $n$}
        \State Sample $x_i^{(k)}$ from $\eta(x_i \mid \vec{x_{i-1}^{(k)}})$
        \Comment Extend the $k$th particle
        \State $w(\vec{x_i}^{(k)}) \gets [\pi(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}) \,/\, \eta(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}})] \cdot w(\vec{x_{i-1}}^{(k)})$
      \EndFor
      \If {$\ESS(w) < T$}
        \Comment{$T$ is a user-defined threshold}
        \State Resample the particles according to $w$
        \For {$k = 1$ to $n$}
          \State $w^{(k)} \gets 1/n$
        \EndFor
      \EndIf
    \EndFor
  \end{algorithmic}
  \label{alg:smc}
\end{algorithm}

{\color{blue}\uline{The resampling step was formally integrated with
\gls{SIS} by \citeauthor{doucet2000sequential} to form the first \gls{SMC}
algorithm }(\cref{alg:smc}). \uline{Rather than resample at every step as the
bootstrap filter proposed, the authors use a criterion based on the \gls{ESS}
the particle population to determine when resampling is necessary. }} The
\gls{ESS} of the population of particles is defined as 
\[ 
  \ESS(w) = \frac{n}{1 + \Var(w)}, 
\]
where $n$ is the number of particles in the population.
When the \gls{ESS} drops below the threshold
{\color{blue}\uline{(conventionally $n/2$~\autocite{liu2008monte})}}, particles
are resampled according to their importance weights. This results in the
removal of low-weight particles from the population, and also equalizes all the
weights. Various resampling strategies beyond the basic sampling with
replacement have been proposed~\autocite{douc2005comparison}, but we will not
discuss those here.

@@

The \gls{SIS} and \gls{SMC} algorithms described above aim to sample from a
high-dimensional distribution $\pi(x)$, by sequentially sampling from $d$
distributions of lower but increasing dimension. \textcite{del2006sequential}
developed an \defn{\gls{SMC} sampler} with an alternative objective: to sample
sequentially from $d$ distributions $\pi_1, \ldots, \pi_d$, all of the same
dimension and defined on the same space. The $\pi_i$ are assumed to form a
related sequence, such as posterior distributions attained by sequentially
considering new evidence. As with \gls{SIS}, we assume that $\pi_i(x) =
\gamma_i(x) / Z_i$, where $\gamma_i$ is known pointwise and the normalizing
constant $Z_i$ is unknown.

@@

\begin{algorithm}
  \caption{Sequential Monte Carlo sampler of \textcite{del2006sequential}.}
  \begin{algorithmic}
    \For {$k = 1$ to $n$}
      \State Sample $x_1^{(k)}$ from $\eta_1(x_1)$
      \Comment{Initialize the $k$th particle}
      \State $w^{(k)} \gets \gamma_1(x_1^{(k)}) / \eta_1(x_1^{(k)})$
      \State Normalize the weights so that $\sum w = 1$
    \EndFor
    \For {$i = 2$ to $d$}
      \For {$k = 1$ to $n$}
        \State Sample $x_i^{(k)}$ from $K(x_{i-1}^{(k)}, x_i)$
        \Comment Extend the $k$th particle
        \State $w^{(k)} \gets w^{(k)} \cdot \dfrac{\gamma_i(x_i) L_{i-1}(x_i, x_{i-1})}{\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}$
      \EndFor
      \State Normalize the weights so that $\sum w = 1$
      \If {$\ESS(w) < T$}
        \Comment{$T$ is a user-defined threshold}
        \State Resample the particles according to $w$
        \For {$k = 1$ to $n$}
          \State $w^{(k)} \gets 1/n$
        \EndFor
      \EndIf
    \EndFor
  \end{algorithmic}
  \label{alg:smcsamp}
\end{algorithm}

@@

For a general model, the probability density of $y$ given the parameters
$\theta$ is also known as the \defn{likelihood}, written $\L$, of $\theta$.
That is, $\L(\theta \mid y) = f(y \mid \theta)$ for the model's \gls{pdf} $f$.
The higher the value of the likelihood, the more likely the observations $y$
are under the model. Thus, the likelihood provides a natural criterion for
fitting the model parameters: we want to pick $\theta$ such that the
probability density of our observed outcomes $y$ is as high as possible. The
parameters which optimize the likelihood are known as the \textit{\gls{ML}}
estimates, denoted $\hat{\theta}$. That is,
\[
  \hat{\theta} = \argmax_\theta\; \L(\theta \mid y).
\]
\Gls{ML} estimation is usually performed with numerical optimization. In the
simplest terms, many possible values for $\theta$ are examined, $\L(\theta \mid
y)$ is calculated for each, and the parameters which produce the highest value
are accepted. Many sophisticated numerical optimization methods exist, although
they may not be guaranteed to find the true \gls{ML} estimates if the
likelihood function is multi-modal.

@@

Ordinarily, Bayesian inference targets the posterior distribution $f(\theta
\mid y)$. That is, in the Bayesian framework, model parameters with higher
posterior density are ``better'' in the sense that they offer a more credible
explanation for the observed data. Approximate Bayesian computation offers an
alternative metric for parameter credibility, namely the similarity of
simulated datasets to the observed data. If datasets simulated under the model
closely resemble the real data, it follows that the model is a reasonable
approximation to the real-world process generating the observed data. More
formally, suppose we have a distance measure $\rho$ defined on the space of all
possible data our model could generate. \gls{ABC} aims to sample from the joint
posterior distribution of model parameters and simulated datasets $z$ which are
within some small distance $\varepsilon$ of the observed data $y$,
\[
  \pi_{\varepsilon}(\theta, z \mid y) =
  \frac{\pi(\theta) f(z \mid \theta) \I_{A_{\varepsilon, y}} (z)}
  {\int_{A_{\varepsilon, y} \times \Theta} \pi(\theta) f(z \mid \theta) \d \theta}.
\]
Here, $A_{\varepsilon, y}$ is an $\varepsilon$-ball around $y$ with
respect to $\rho$, $\Theta$ is the space of all possible model parameters, and
$\I$ is the indicator function~\autocite{marin2012approximate}. As we shall
see in the next section, this distribution can be sampled from exactly. The
word ``approximate'' derives from the assumption that, for a suitably chosen
distance $\rho$ and a small enough $\varepsilon$, the marginal in $z$ of
this distribution approximates the posterior of
interest~\autocite{marin2012approximate}. That is,
\[
  \int \pi_\varepsilon(\theta, z \mid y) \d z \approx f(\theta \mid y).
\]
The distribution $\pi_\varepsilon(\theta, z \mid y)$ is variously referred to
as the \textit{\gls{ABC} target distribution} or the \gls{ABC} approximation
to the posterior. The intuition for why this approximation might be reasonable
comes from the fact that, when $\varepsilon = 0$ and $\rho$ satisfies $\rho(z,
y) = 0 \Rightarrow x = y$, the integral on the left is exactly equal to the
posterior. Thus, by taking $\varepsilon$ small, we should attain something
close to the posterior {\color{blue}\uline{if $\rho$ captures the similarity
between datasets reasonably well. However, the accuracy of the \gls{ABC}
approximation depends heavily on the choice of distance function. }}

In many applications {\color{purple}\uline{(eg. \autocite{fu1997estimating,
tanaka2006using})}}, $\rho$ is defined as $\rho(S(\cdot), S(\cdot))$ where $S$
is a function which maps data points into a vector of summary statistics.
{\color{blue}\uline{In the context of \gls{ABC}, a summary statistic $S$ is
called \defn{sufficient} if} 
\[
  f(\theta \mid y) = f(\theta \mid S(y)).
\]
\uline{That is, sufficiency implies that the data can be replaced with the
summary statistic without losing any information about the posterior
distribution~\autocite{marjoram2006modern}. For most problems, it is not
possible to find sufficient summary statistics~\autocite{marjoram2006modern}.
TODO
}}

This can be useful if the data are high-dimensional or of a complex type, but
it is not strictly necessary. For instance, if the data are numeric and of low
dimension, the distance function may simply be the Euclidean
distance~\autocite{sisson2007sequential}. \textcite{poon2015phylodynamic}
proposed the use of a kernel function (defined in \cref{subsec:treeshape}),
which allows

@@

Algorithms for performing \gls{ABC} fall into one of three categories:
rejection, \gls{MCMC}, and \gls{SMC}~\autocite{marin2012approximate}. To
simplify the notation, we shall restrict the descriptions of these algorithms
to the case of one simulated dataset per parameter particle (the meaning of
this will become clear shortly). The extension to multiple datasets per
particle is straightforward and will be given at the end of the section. We use
the variable $x$ to refer to the pair $(\theta, z)$, so that the \gls{ABC}
target distribution can be written $\pi_\varepsilon(x \mid y)$.
